Density-dependent selection and the limits of relative fitness

### 2017-12-12 manuscript.tex
l

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

@@

Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra mode, see \citep{smouse_1976,mallet_2012}).

>>>>

It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

@@

Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra model, see \citep{smouse_1976,mallet_2012}).
